A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-08-01 papers/water-saft/paper.tex
modified plots, added citations, corrected hydrogen bond breaking calculations in water-saft Tested everything on node09.

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)~\fixme{add citations here,
  particularly Jeanmairet 2013 and his citations 35-44, possibly
  cutting non-acqueous papers from that list}.
Classical DFT is based on a description of a fluid written as a free
energy functional of the density distribution.  There are two
approaches used to construct a classical DFT for water.  The first is
to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure~\fixme{cite Jeanmairet and
  similar}.  Using this approach, it is possible to construct a
functional that reproduces the exact second-order response function of
the liquid under the fitted conditions.  This class of functional,
however, will be less accurate at other temperatures or
pressures---and in the inhomogeneous scenarios in which solvation
models are applied.  The second approach is construct a functional by
applying liquid-state theory to a model system, and then fit the model
to experimental data such as the equation of state.

@@

We modify our previous the association free energy term to use our
$g_{\sigma}^A$ rather than Yu and Wu's $g_\sigma^{YuWu}$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{previous paper}.
  \end{cases} 
\end{align}
The four in $n_{site}(\rr)$ comes from the four hydrogen bond sites
and, $n_o(\rr)$ is the average density at contact, and $\zeta(\xx)$ is
a dimensionless measure of the density inhomogeneity from Yu and
Wu. \cite{yu2002fmt-dft-inhomogeneous-associating} The functional
$X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of hydrophobic rods of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}^{HS}$.}
\label{fig:single-rod-X}
\end{figure}

@@

\section{Introduction}

>>>>

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}
~\fixme{add citations here, particularly Jeanmairet 2013 and his
  citations 35-44, possibly cutting non-acqueous papers from that
  list}.  Classical DFT is based on a description of a fluid written
as a free energy functional of the density distribution.  There are
two approaches used to construct a classical DFT for water.  The first
is to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.

@@

We modify our previous the association free energy term to use our
$g_{\sigma}^A$ rather than Yu and Wu's $g_\sigma^{YuWu}$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites at position~$\rr$.  The four in
$n_\text{site}(\rr)$ comes from the four hydrogen bond sites,
$n_0(\rr)$ is the average density contacting point $\rr$, and
$\zeta(\xx)$ is a dimensionless measure of the density inhomogeneity
from Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}.  The
functional $X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 2n_\text{site}'(\rr)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {n_\text{site}'(\rr)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  n_\text{site}'(\rr) &=
  \begin{cases}
    \frac{4}{\pi\sigma^2} \int n(\rr')\delta(\sigma - |\rr-\rr'|) d\rr' & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites that could bond to the sites~$n_\text{site}(\rr)$, and
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of hydrophobic rods of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}^{HS}$.}
\label{fig:single-rod-X}
\end{figure}
\section{Introduction}
